A Multi-Modal Model of Object Deformation under Robotic Pushing

### 2017-01-24 twostagedo-rev-final.tex
Added source of final version.

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
In this paper we present a multi-modal framework for \comment{offline} learning of generative models of object deformation under robotic pushing. The model is multi-modal in that it is based on integrating force and visual information. The framework consists of several sub-models that are independently \comment{calibrated} from the same data. These component models can be sequenced to provide many-step prediction and classification. When presented with a test example--a robot finger pushing a deformable object made of an unidentified, but previously learned, material--the predictions of modules for different materials are compared so as to classify the unknown material. Our approach, which consists of \comment{offline learning and combination of} multiple models, goes beyond previous techniques by enabling i) predictions over many steps, ii) learning of plastic and elastic deformation from real data, iii) prediction of forces experienced by the robot, iv) classification of materials from both force and visual data, v) prediction of object behaviour after contact by the robot terminates.  While previous work on deformable object behaviour in robotics has offered one or two of these features none has offered a way to achieve them all, and none has offered classification from a generative model. We do so through separately learned models which can be combined in different ways for different purposes.
\end{abstract}

@@

There are various models of deformation used in engineering, such as finite element models, but these typically require careful optimisation by hand, and cannot run in real time. In graphics there are many methods used for generating qualitatively plausible simulations of deformable objects, but these are not easy to fit to real data. \comment{In contrast, robots must learn model parameters from data, in reasonable amounts of training time. This includes learning from the limited sensing typically available, such as training data gathered from a specific view(s) of an object, and force-torque monitoring only at contact points.} Thus, tackling the prediction problem in robotics adds additional difficulties to an already challenging problem. \comment{In this paper we address some of these---we learn from real visual and force-torque data---while simplifying others---we set a convenient fixed viewing location and perform off-line rather than on-line learning.}

@@

\comment{There are two novel technical contributions. First, we show how the learning problem can be decomposed into two separate learning problems, one for predicting forces from finger motions, and one for predicting object deformations from forces. Both models are generative, and we show how to learn each from data. Second, we show how to use these learned generative models not only to predict deformation, but also to classify an unidentified material. To limit the scope of the paper we develop this framework using offline learning methods. We do not address the issue of online learning \cite{worgotter09,antonelli14}, leaving this for future work.}

@@

\comment{These technical contributions} enable us to learn models of sensorimotor contingencies, and then to use these as components to build different machines. This approach enables i) predictions over many steps, ii) learning of plastic and elastic deformation from real data, iii) prediction of forces experienced by the robot, iv) classification of materials from either or both force and visual data, v) prediction of object behaviour after contact by the robot is removed. In the rest of the paper we present related work (Section~\ref{sec:background}); give an overview of the framework (Section~\ref{sec:model_overview}); describe the two component models in detail (Section~\ref{sec:model}); describe how the model parameters are learned from real data for the shape prediction model (Section~\ref{sec:learning}) and for the force prediction model (Section~\ref{sec:forces}); report experiments on prediction and classification with elastic and plastic objects (Section~\ref{sec:experiments}); and finish with a short discussion (Section~\ref{sec:discussion}).

@@

Teschner et al.\ \cite{Teschner2004} proposed an enhanced 3D mass-spring model, in which constraints are expressed as potential energy terms.  These terms encode the tendency of springs to recover their length (with damping), of triangular faces of a mesh to recover their area and of tetrahedrons to recover their volume.  He sketches the inclusion of plastic deformation.  Morris and Salisbury \cite{Morris2008} developed a method to calibrate Teschner's model automatically with respect to a \comment{finite element model (FEM)}.  The  search begins with a uniform random sampling of possible values for the elasticity constants, which are latter modified using adaptive simulated annealing to minimize the difference between the deformations in the FEM mesh and those in the mass-spring mesh.  However, they only compare the steady states and do not include plastic deformations.  We extend this work, by replacing the FEM model with force and 2D visual data gathered from real materials.  Furthermore, we analyse the quality of the predictions while the deformations take place.

@@

\begin{figure}[!t]
\centering
%\setlength\fboxsep{0.5pt}    % padding
%\setlength\fboxrule{0.5pt} % line width
%\fbox{
\includegraphics[width=3.5in]{figures/learning.png}%}
\caption{\comment{The visual tracking and force measurement system subserves the offline learning system. The offline learning (or calibration) method comprises two learned models of sensorimotor contingencies.} Both are learned from the same data. The first (FP) learns to predict the force experienced for some motion and material. The second (SP) predicts the deformation of the object given the applied force. The strain $st$ is calculated directly from the interpenetration of the finger into the object. This can be calculated by vision (V) or from the planned motion of the finger.}
\label{fig:learning}
\end{figure}

@@

\comment{The framework consists of four separate systems: a visual tracking and force measurement system, an offline learning (or model calibration) system, a prediction system, and a classification system.

Visual tracking and force measurement both serve the learning system (Figure~\ref{fig:learning}). The visual analysis is not a contribution of the paper, any method that provides accurate tracking of the outline of the deformed object would suffice. We employ a Canny edge detector, filtering out irrelevant edges using the colour of the manipulated object. Then we fit the outline and track using a linear snake. This is also used to initialise the mesh for the spring mass system.}

@@

\comment{In machine learning there is division between discriminative and generative models. Discriminative models are powerful for classification, but cannot be used to regenerate the data, or to generate new imagined data. In our application this would mean we could not predict deformations, only being able to classify materials directly based on their deformations. Instead, following others, we use a generative model. But we show how this can be used not just for prediction, but also for material classification. Thus a generative model is appealing because it allows us to solve both the prediction and classification problems.}

@@

\subsection{Tracking to Provide Ground Truth}
\begin{figure}[!t]
 \begin{algorithmicieee}{Regularised Linear Spline}\label{alg:regularise}
 \STATE Let $L$ be a linear spline representing the contour of a $2D$ shape, and $L_i$ the $i^{th}$ control point in a cyclic order.
 \FORALL {$L_i$ in $L$}
 \IF {$dist(L_i,L_{i-1}) > max\_distance$}
  \STATE Insert $midpoint(L_i,L_{i-1})$ after $L_i$
 \ELSIF {$dist(L_i,L_{i-1}) < min\_distance$}
  \STATE erase $L_i$
 \ELSIF {$angle(L_{i-1},L_i,L_{i+1}) < min\_angle$}
  \STATE erase $L_i$
 \ENDIF
 \ENDFOR
\end{algorithmicieee}
\caption{Algorithm used to regularise the number of elements of a linear spline, in accordance with the level of detail required.}
\end{figure}
 The contour of the deformable object is used to obtain the ground truth for the training phase, which is the area enclosed by the contour. Since we only require a tracking algorithm which is good enough to evaluate the predictions of the mass-spring system, a very simple one was used.  Geometrically, the contour is represented by a polygon with hundreds of sides, which can also be called a \textit{linear spline} \cite{Menet1991}. The vertices of the polygon work as control points which can be moved to adjust the polygon to the shape of the object as it gets deformed.  \comment{We call our algorithm for tracking a \emph{linear snake}, because}  this implementation is inspired by the theory of active contours, however it works with a much simpler and faster, but less accurate, algorithm, instead of minimizing a global energy.  \comment{This causes some shadowed corners of the material to be missed at some frames, but as \fref{fig:scene}(b) at its bottom left shows, these errors are not significant in comparison to the errors made by the mass-spring model.} This is explained below.  The tracking of the deformable target in the ground truth has two stages:
 \begin{figure}[t]
\centering
\includegraphics[width=88mm]{arrio2}
\caption{(a) Experimental set up.  A robot finger pushes a block of deformable material sitting on a table.  A camera registers the events of the plane where the main deformations take place.  (b) Evaluation function.  Left top: Mesh over real block of material.  Left bottom: The typical error of the model vs. the ground truth (red) and the typical error of the tracker (blue).  Right top: tracked ground data.  Middle: TP in yellow, FP in blue, FN in red and TN in gray.  Bottom: simulated mesh. }
\label{fig:scene}
\end{figure}

\begin{enumerate}
 \item The target object is segmented from the environment.  The Canny algorithm is used to extract the edges of the image.  Since the object of interest has a distinctive colour, \comment{its characteristic hue range is extracted from the area enclosed by the initial contour.  Then} the hue of the pixels on every edge are used to keep only edges that could belong to it.  If the average hue of the pixels on an edge falls outside the calibrated range, the edge is eliminated.
 \item The internal representation of its contour, the linear snake, is adapted to its new shape.  The algorithm assumes that no occlusions take place and that the contour does not bend over itself.
\end{enumerate}
For the first frame, the snake is initialised as a rectangle around the target. This rectangle fragments itself into smaller pieces. Each new vertex adheres itself to the closest point on an edge, as reached by growing rings of pixels around its current position, up to a certain threshold distance [\fref{fig:scene}(a)].  If there is no edge within that area, the control point remains where it is. This criterion was selected to make the representation more robust against edges that are easier to detect on some frames than in others.  Once the edge reappears, the control point adapts itself again.  Also, the distance between two consecutive control points is kept within the interval $[min\_distance, max\_distance]$ and two adjacent edges do not form angles smaller than $min\_angle$.  See \alref{alg:regularise}.  This allows the snake to adapt very quickly to the complexity of the contour it represents.

@@

When one or more particles are displaced from their equilibirum positions, the forces that tend to restore an elastic body to its original shape are modeled by the forces of springs (along the edges of the mesh) that try to recover their original length).  These forces act upon the particles modifying their positions in time.  Traditionally, the relationship between the position of the particle $p_i$ and those forces $F$ obeys the second law of Newton:
\begin{align}
F = m \dfrac{\partial \vec{p}}{\partial t} \label{eq:newton}
\end{align}
where $t$ is time.  However, it is possible to deviate from this equation if we need to model a different behaviour. %are interested in is numerical accuracy, as it will be done here.

@@

To simulate the effect of an external actuator (like the finger of the robot) interacting with the deformable object, external forces $F_{ext}$ acting on vertices of the mesh must be added to \eref{eq:newton}:
\begin{align}
F + F_{ext} = m \dfrac{\partial \vec{p}}{\partial t}
\end{align}
Another option could be to induce response forces by displacing vertices from their position of equilibrium.

@@

The mass-spring model proposed by Teschner et al.\ in \cite{Teschner2004}, which is explained below, was reduced to a 2D model, so that the predictions could be compared with 2D images of the deformable object.  The main idea being that an agent can make rough real time predictions of what it can see and touch and use those predictions to solve various tasks, like classification or grasping.  Here, the surface of the deformable object is approximated by a regular triangular mesh of unitary masses and springs.  \comment{Since the location of the masses and direction of the springs will bias the deformation response of the object, it is important to distribute the masses uniformly and place the springs with a symetric arrangement for an isotropic material.  This placement can be done with an algorithm that takes into account the symmetries of the object at the begining of the experiment.  If the material gets permanently deformed a properly calibrated physics based model, like the one we propose here, will produce an adequate new arrangement from which new interactions can be studied.}

@@

\subsubsection{Preservation of Angles}
The previous terms can not do anything to restore the mesh if triangles get flipped during deformation.  For this reason we included a new term that enforces the preservation of the original angles of the triangles, \comment{which is not a valid state since the mesh models the surface of a 3D object}.  [\fref{fig:forces}(c)] gives an idea of how these forces look.  Here the energy depends on the difference \comment{%\sout{of the angles between adjacent edges.} 
between the current angle, between adjacent edges, and the angle between them at the equilibrium position.} Energy is given by\footnote{It was also considered to multiply $E_\varphi$ by the lengths of the edges, but it hasn't improved the performance of the model.}:
\begin{align}
 E_\varphi(\varphi) & = \frac{1}{2}k_\varphi(\varphi - \varphi_0)^2 \\
 \varphi(p_i,p_j,p_k) & = \arccos \left( \dfrac{(p_j - p_i) \cdot (p_k - p_i)}{ \left\| p_j - p_i \right\| \left\|p_k - p_i \right\|} \right)^2 \nonumber
\end{align}

@@

\subsection{Integration Scheme}
\label{sec:ischeme}
Rather than using traditional Newtonian equations with a numeric integration scheme like Euler or Verlet, used twice to solve the second order differential equiations, we made the force proportional to the velocity:
\comment{
\begin{align}
 F(t) = \vec{p}
\end{align}
which implies the following assumption:
\begin{align}
 \dfrac{\partial \vec{p}}{\partial t} = \vec{p}
\end{align}
this equation can only be satisfied with an exponential function, thus avoiding the oscillatory behaviour of sinusoid solutions for springs, in favour of forcing critically damped springs to model the elastic materials.
}

@@

\subsection{Additional Geometric Constraints}
While the mass-spring model incorporates constraints based on mesh geometry, these are soft constraints implemented in the form of an energy function. Even with several of these terms incorporated, such as a term for preservation of angles, the mass-spring model can cause triangles of the mesh to overlap one another and be reversed or flattened.  \comment{All of these situations are inadequate since this mesh is used to model one face of a 3D object.}  It is thus necessary to employ additional hard geometric constraints to make the mesh internally consistent.  We achieve this through geometric checking subroutines.  There are two geometric constraints that the mesh is enforced to maintain:

\begin{enumerate}
 \item For a reversed triangle, (one vertex crossed over an edge): the vertex gets pushed more than half the distance between the vertex and the edge in the direction of the ``reversed'' height\footnote{$3/4$ of this distance was used in our implementation.}, while both ends of the edge get pushed the same amount in the opposite direction.  This is enough for most cases, but triangles in the border of the mesh can still overlap each other \comment{which would render a bad numerical approximation to the behaviour of a surface}.

@@

The evolutionary algorithm described in \alref{alg:genetic} was used to search the parameter space of the mass-spring model presented in the previous section.
From the parameters that must be chosen for the simulation to work, some were fixed while others had to be searched for.  They are:
\begin{itemize}
 \item The mass per vertex.  \textit{(For the moment we consider a unitary mass for all vertices.)}
 \item The elasticity constants: for the preservation of length $k_L$, area $k_A$, angles $k_{\varphi}$ and linear damping $k_d$.  \textit{(All springs will have the same values to favour generalisation.)}
 \item The plastic parameters: $yield$, $creep$ and the maximum amount of permanent deformation $max\_\alpha$.  \textit{(All springs will have the same values to favour generalisation.)}
  \item The integration time step $\Delta t$.  \textit{(It has been set to 0.1 seconds, even though values as small as 0.01 were tried as well; \comment{however, since they didn't improve stability or accuracy in a visible way, but made the process much slower, they were disregarded as impractical}.)}
 \item An upper limit for the magnitude of the total estimated forces per vertex $max\_Force$.
 \item The minimum area before a triangle is considered \textit{flat}.  (It has been set to $2.0$ units.)
\end{itemize}
To evaluate the suitability of each set of parameters, the simulation is compared with the ground truth tracked from real objects.

@@

\section{Experimental Results}
\label{sec:experiments}
\subsection{Scenario}
For the experimental scenario we used two materials: sponge and plasticine. Each formed a rectangular block of deformable material, of $8.5 \unit{cm}$ long, $5 \unit{cm}$ wide and $1.8 \unit{cm}$ depth. One of the objects is selected, and is placed on a table; the robot pushes the object with a cylindrical finger, followed by a retrieval movement.  A colour firewire camera records the action perpendicularly, while a force sensor registers the reaction forces in synchrony with each photograph taken [\fref{fig:scene}(a)].  \comment{This set up allowed us to study the effect of using an automatically calibrated physics based model as model for the material, while using 2D meshes, even though the same principles can be used if we use a 3D point cloud and a 3D mesh.  To generate the initial mesh for the mass-spring model a routine placed the nodes for the masses evenly spaced on a rectangular grid, given the number of squares we wanted per side.  We made experiments with grids of $3 \times 2, 10 \times 5$ and $20 \times 10$ squares.  To alleviate the bias produced by the springs they were placed in a star pattern, thus compensating forces on eight directions.}

@@

The traditional equations of movement, estimated with Euler or Verlet show big oscillations and frequently become unstable.  Verlet seems to be badly affected by the collision resolution routines and geometric aids, that drastically affect the values of velocity and acceleration.  On the other hand, the customised first degree equations suggested in \sref{sec:ischeme} produce smooth movements that more closely resemble the ground truth, thus obtaining better marks.  From this group of experiments it was also found that evolutionary learning is as efficient when considering all elastic and plastic terms, as when using only a few of them\footnote{\comment{The full list of terms is summarised at the end of \sref{sub:EvolAlg}.}}.

\subsection{Results of Automatic Parameter Calibration}
\subsubsection{Performance for the Learned Shape Prediction Model}
The performance of the genetic learning algorithm was evaluated using the customised integration scheme and the mass-spring elastic-plastic model.  At this point no prediction of the force was included, rather, the reaction forces measured with the force sensor were used. \comment{The F-score was used as fitness function. The values used for the parameters of \alref{alg:genetic} are summarised in \tref{tab:gen_params}, also the minimum and maximum limits for the values of parameters for the mass-spring model.}

\begin{table}[!t]
 \renewcommand{\arraystretch}{1.3}
 \caption{Parameters for the Genetic Algorithm}
 \label{tab:gen_params}
 \centering
\comment{
\begin{tabular}{cc}
\hline
 \bfseries Parameter & \bfseries Value/Range \\  \hline
 \multicolumn{2}{c}{Genetic Algorithm} \\ \hline
 M & 30 \\
 $\alpha = \beta = \gamma = \zeta$  & 5 \\
 $\sigma_{max}$                     & 1000.0 \\
 $\sigma_{min}$                     & 10.0 \\
 $P(s)$                             & 0.05 \\ \hline
 \multicolumn{2}{c}{Elastic Parameters } \\ \hline
 $k_d$                              & $[0.00001, 2]$ \\
 $K_L$                              & $[0.00001, 10000]$ \\
 $K_A$                              & $[0.00001, 10000]$ \\
 $K_{\varphi}$                      & $[0.00001, 100]$ \\
 $max(F)$                           & $[5, 100000]$ \\ \hline
 \multicolumn{2}{c}{Plastic Parameters } \\ \hline
 $\alpha$                           & $[0.00001, 0.9]$ \\
 $yield$                            & $[0.00001, 1.0]$ \\
 $creep$                            & $[0.00001, 1.0]$ \\ \hline
\end{tabular}

\vspace*{1em}
* Where $P(s)$ is the probability of using a special value, like $0.0$, for the spring constants.
}
\end{table}

>>>>

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
In this paper we present a multi-modal framework for offline learning of generative models of object deformation under robotic pushing. The model is multi-modal in that it is based on integrating force and visual information. The framework consists of several sub-models that are independently calibrated from the same data. These component models can be sequenced to provide many-step prediction and classification. When presented with a test example--a robot finger pushing a deformable object made of an unidentified, but previously learned, material--the predictions of modules for different materials are compared so as to classify the unknown material. Our approach, which consists of offline learning and combination of multiple models, goes beyond previous techniques by enabling i) predictions over many steps, ii) learning of plastic and elastic deformation from real data, iii) prediction of forces experienced by the robot, iv) classification of materials from both force and visual data, v) prediction of object behaviour after contact by the robot terminates.  While previous work on deformable object behaviour in robotics has offered one or two of these features none has offered a way to achieve them all, and none has offered classification from a generative model. We do so through separately learned models which can be combined in different ways for different purposes.
\end{abstract}

@@

There are various models of deformation used in engineering, such as finite element models, but these typically require careful optimisation by hand, and cannot run in real time. In graphics there are many methods used for generating qualitatively plausible simulations of deformable objects, but these are not easy to fit to real data. In contrast, robots must learn model parameters from data, in reasonable amounts of training time. This includes learning from the limited sensing typically available, such as training data gathered from a specific view(s) of an object, and force-torque monitoring only at contact points. Thus, tackling the prediction problem in robotics adds additional difficulties to an already challenging problem. In this paper we address some of these---we learn from real visual and force-torque data---while simplifying others---we set a convenient fixed viewing location and perform off-line rather than on-line learning.

@@

There are two novel technical contributions. First, we show how the learning problem can be decomposed into two separate learning problems, one for predicting forces from finger motions, and one for predicting object deformations from forces. Both models are generative, and we show how to learn each from data. Second, we show how to use these learned generative models not only to predict deformation, but also to classify an unidentified material. To limit the scope of the paper we develop this framework using offline learning methods. We do not address the issue of online learning \cite{worgotter09,antonelli14}, leaving this for future work.

@@

These technical contributions enable us to learn models of sensorimotor contingencies, and then to use these as components to build different machines. This approach enables i) predictions over many steps, ii) learning of plastic and elastic deformation from real data, iii) prediction of forces experienced by the robot, iv) classification of materials from either or both force and visual data, v) prediction of object behaviour after contact by the robot is removed. In the rest of the paper we present related work (Section~\ref{sec:background}); give an overview of the framework (Section~\ref{sec:model_overview}); describe the two component models in detail (Section~\ref{sec:model}); describe how the model parameters are learned from real data for the shape prediction model (Section~\ref{sec:learning}) and for the force prediction model (Section~\ref{sec:forces}); report experiments on prediction and classification with elastic and plastic objects (Section~\ref{sec:experiments}); and finish with a short discussion (Section~\ref{sec:discussion}).

@@

Teschner et al.\ \cite{Teschner2004} proposed an enhanced 3D mass-spring model, in which constraints are expressed as potential energy terms.  These terms encode the tendency of springs to recover their length (with damping), of triangular faces of a mesh to recover their area and of tetrahedrons to recover their volume.  He sketches the inclusion of plastic deformation.  Morris and Salisbury \cite{Morris2008} developed a method to calibrate Teschner's model automatically with respect to a finite element model (FEM).  The  search begins with a uniform random sampling of possible values for the elasticity constants, which are latter modified using adaptive simulated annealing to minimize the difference between the deformations in the FEM mesh and those in the mass-spring mesh.  However, they only compare the steady states and do not include plastic deformations.  We extend this work, by replacing the FEM model with force and 2D visual data gathered from real materials.  Furthermore, we analyse the quality of the predictions while the deformations take place.

@@

\begin{figure}[!t]
\centering
%\setlength\fboxsep{0.5pt}    % padding
%\setlength\fboxrule{0.5pt} % line width
%\fbox{
\includegraphics[width=3.5in]{figures/learning.png}%}
\caption{The visual tracking and force measurement system subserves the offline learning system. The offline learning (or calibration) method comprises two learned models of sensorimotor contingencies. Both are learned from the same data. The first (FP) learns to predict the force experienced for some motion and material. The second (SP) predicts the deformation of the object given the applied force. The strain $st$ is calculated directly from the interpenetration of the finger into the object. This can be calculated by vision (V) or from the planned motion of the finger.}
\label{fig:learning}
\end{figure}

@@

The framework consists of four separate systems: a visual tracking and force measurement system, an offline learning (or model calibration) system, a prediction system, and a classification system.

Visual tracking and force measurement both serve the learning system (Figure~\ref{fig:learning}). The visual analysis is not a contribution of the paper, any method that provides accurate tracking of the outline of the deformed object would suffice. We employ a Canny edge detector, filtering out irrelevant edges using the colour of the manipulated object. Then we fit the outline and track using a linear snake. This is also used to initialise the mesh for the spring mass system.

@@

In machine learning there is division between discriminative and generative models. Discriminative models are powerful for classification, but cannot be used to regenerate the data, or to generate new imagined data. In our application this would mean we could not predict deformations, only being able to classify materials directly based on their deformations. Instead, following others, we use a generative model. But we show how this can be used not just for prediction, but also for material classification. Thus a generative model is appealing because it allows us to solve both the prediction and classification problems.

@@

\subsection{Tracking to Provide Ground Truth}
\begin{figure}[!t]
 \begin{algorithmicieee}{Regularised Linear Spline}\label{alg:regularise}
 \STATE Let $L$ be a linear spline representing the contour of a $2D$ shape, and $L_i$ the $i^{th}$ control point in a cyclic order.
 \FORALL {$L_i$ in $L$}
 \IF {$dist(L_i,L_{i-1}) > max\_distance$}
  \STATE Insert $midpoint(L_i,L_{i-1})$ after $L_i$
 \ELSIF {$dist(L_i,L_{i-1}) < min\_distance$}
  \STATE erase $L_i$
 \ELSIF {$angle(L_{i-1},L_i,L_{i+1}) < min\_angle$}
  \STATE erase $L_i$
 \ENDIF
 \ENDFOR
\end{algorithmicieee}
\caption{Algorithm used to regularise the number of elements of a linear spline, in accordance with the level of detail required.}
\end{figure}
 The contour of the deformable object is used to obtain the ground truth for the training phase, which is the area enclosed by the contour. Since we only require a tracking algorithm which is good enough to evaluate the predictions of the mass-spring system, a very simple one was used.  Geometrically, the contour is represented by a polygon with hundreds of sides, which can also be called a \textit{linear spline} \cite{Menet1991}. The vertices of the polygon work as control points which can be moved to adjust the polygon to the shape of the object as it gets deformed.  We call our algorithm for tracking a \emph{linear snake}, because this implementation is inspired by the theory of active contours, however it works with a much simpler and faster, but less accurate, algorithm, instead of minimizing a global energy.  This causes some shadowed corners of the material to be missed at some frames, but as \fref{fig:scene}(b) at its bottom left shows, these errors are not significant in comparison to the errors made by the mass-spring model.  This is explained below.  The tracking of the deformable target in the ground truth has two stages:
 \begin{figure}[t]
\centering
\includegraphics[width=88mm]{arrio2}
\caption{(a) Experimental set up.  A robot finger pushes a block of deformable material sitting on a table.  A camera registers the events of the plane where the main deformations take place.  (b) Evaluation function.  Left top: Mesh over real block of material.  Left bottom: The typical error of the model vs. the ground truth (red) and the typical error of the tracker (blue).  Right top: tracked ground data.  Middle: TP in yellow, FP in blue, FN in red and TN in gray.  Bottom: simulated mesh. }
\label{fig:scene}
\end{figure}

\begin{enumerate}
 \item The target object is segmented from the environment.  The Canny algorithm is used to extract the edges of the image.  Since the object of interest has a distinctive colour, its characteristic hue range is extracted from the area enclosed by the initial contour.  Then the hue of the pixels on every edge are used to keep only edges that could belong to it.  If the average hue of the pixels on an edge falls outside the calibrated range, the edge is eliminated.
 \item The internal representation of its contour, the linear snake, is adapted to its new shape.  The algorithm assumes that no occlusions take place and that the contour does not bend over itself.
\end{enumerate}
For the first frame, the snake is initialised as a rectangle around the target. This rectangle fragments itself into smaller pieces. Each new vertex adheres itself to the closest point on an edge, as reached by growing rings of pixels around its current position, up to a certain threshold distance [\fref{fig:scene}(a)].  If there is no edge within that area, the control point remains where it is. This criterion was selected to make the representation more robust against edges that are easier to detect on some frames than in others.  Once the edge reappears, the control point adapts itself again.  Also, the distance between two consecutive control points is kept within the interval $[min\_distance, max\_distance]$ and two adjacent edges do not form angles smaller than $min\_angle$.  See \alref{alg:regularise}.  This allows the snake to adapt very quickly to the complexity of the contour it represents.

@@

When one or more particles are displaced from their equilibirum positions, the forces that tend to restore an elastic body to its original shape are modeled by the forces of springs (along the edges of the mesh) that try to recover their original length).  These forces act upon the particles modifying their positions in time.  Traditionally, the relationship between the position of the particle $p_i$ and those forces $F$ obeys the second law of Newton:
\begin{align}
F = m \dfrac{\partial^2 \vec{p}}{\partial t^2} \label{eq:newton}
\end{align}
where $t$ is time and $\vec{p}$, the position.  However, it is possible to deviate from this equation if we need to model a different behaviour. %are interested in is numerical accuracy, as it will be done here.

@@

To simulate the effect of an external actuator (like the finger of the robot) interacting with the deformable object, external forces $F_{ext}$ acting on vertices of the mesh must be added to \eref{eq:newton}:
\begin{align}
F + F_{ext} = m \dfrac{\partial^2 \vec{p}}{\partial t^2}
\end{align}
Another option could be to induce response forces by displacing vertices from their position of equilibrium.

@@

The mass-spring model proposed by Teschner et al.\ in \cite{Teschner2004}, which is explained below, was reduced to a 2D model, so that the predictions could be compared with 2D images of the deformable object.  The main idea being that an agent can make rough real time predictions of what it can see and touch and use those predictions to solve various tasks, like classification or grasping.  Here, the surface of the deformable object is approximated by a regular triangular mesh of unitary masses and springs.  Since the location of the masses and direction of the springs will bias the deformation response of the object, it is important to distribute the masses uniformly and place the springs with a symetric arrangement for an isotropic material.  This placement can be done with an algorithm that takes into account the symmetries of the object at the begining of the experiment.  If the material gets permanently deformed a properly calibrated physics based model, like the one we propose here, will produce an adequate new arrangement from which new interactions can be studied.

@@

\subsubsection{Preservation of Angles}
The previous terms can not do anything to restore the mesh if triangles get flipped during deformation, which is not a valid state since the mesh models the surface of a 3D object.  For this reason we included a new term that enforces the preservation of the original angles of the triangles.  [\fref{fig:forces}(c)] gives an idea of how these forces look.  Here the energy depends on the difference between the current angle, between adjacent edges, and the angle between them at the equilibrium position. Energy is given by\footnote{It was also considered to multiply $E_\varphi$ by the lengths of the edges, but it hasn't improved the performance of the model.}:
\begin{align}
 E_\varphi(\varphi) & = \frac{1}{2}k_\varphi(\varphi - \varphi_0)^2 \\
 \varphi(p_i,p_j,p_k) & = \arccos \left( \dfrac{(p_j - p_i) \cdot (p_k - p_i)}{ \left\| p_j - p_i \right\| \left\|p_k - p_i \right\|} \right)^2 \nonumber
\end{align}

@@

\subsection{Integration Scheme}
\label{sec:ischeme}
Rather than using traditional Newtonian equations with a numeric integration scheme like Euler or Verlet, used twice to solve the second order differential equations, we made the force proportional to the velocity:
\begin{align}
 F(t) = m\dfrac{\partial \vec{p}(t)}{\partial t} = m \vec{v}(t)
\end{align}
which implies the following assumption:
\begin{align}
 \vec{a}(t) = \dfrac{\partial \vec{v}(t)}{\partial t} = \vec{v}(t)
\end{align}
this equation of the form $f=f'$ can only be satisfied with an exponential function $f(x)=e^x$, thus avoiding the oscillatory behaviour of sinusoid solutions for springs, in favour of forcing critically damped springs to model the elastic materials.

@@

\subsection{Additional Geometric Constraints}
While the mass-spring model incorporates constraints based on mesh geometry, these are soft constraints implemented in the form of an energy function. Even with several of these terms incorporated, such as a term for preservation of angles, the mass-spring model can cause triangles of the mesh to overlap one another and be reversed or flattened.  All of these situations are inadequate since this mesh is used to model one face of a 3D object.  It is thus necessary to employ additional hard geometric constraints to make the mesh internally consistent.  We achieve this through geometric checking subroutines.  There are two geometric constraints that the mesh is enforced to maintain:

\begin{enumerate}
 \item For a reversed triangle, (one vertex crossed over an edge): the vertex gets pushed more than half the distance between the vertex and the edge in the direction of the ``reversed'' height\footnote{$3/4$ of this distance was used in our implementation.}, while both ends of the edge get pushed the same amount in the opposite direction.  This is enough for most cases, but triangles in the border of the mesh can still overlap each other which would render a bad numerical approximation to the behaviour of a surface.

@@

The evolutionary algorithm described in \alref{alg:genetic} was used to search the parameter space of the mass-spring model presented in the previous section.
From the parameters that must be chosen for the simulation to work, some were fixed while others had to be searched for.  They are:
\begin{itemize}
 \item The mass per vertex.  \textit{(For the moment we consider a unitary mass for all vertices.)}
 \item The elasticity constants: for the preservation of length $k_L$, area $k_A$, angles $k_{\varphi}$ and linear damping $k_d$.  \textit{(All springs will have the same values to favour generalisation.)}
 \item The plastic parameters: $yield$, $creep$ and the maximum amount of permanent deformation $max\_\alpha$.  \textit{(All springs will have the same values to favour generalisation.)}
  \item The integration time step $\Delta t$.  \textit{(It has been set to 0.1 seconds, even though values as small as 0.01 were tried as well; however, since they didn't improve stability or accuracy in a visible way, but made the process much slower, they were disregarded as impractical.)}
 \item An upper limit for the magnitude of the total estimated forces per vertex $max\_Force$.
 \item The minimum area before a triangle is considered \textit{flat}.  (It has been set to $2.0$ units.)
\end{itemize}
To evaluate the suitability of each set of parameters, the simulation is compared with the ground truth tracked from real objects.

@@

\section{Experimental Results}
\label{sec:experiments}
\subsection{Scenario}
For the experimental scenario we used two materials: sponge and plasticine. Each formed a rectangular block of deformable material, of $8.5 \unit{cm}$ long, $5 \unit{cm}$ wide and $1.8 \unit{cm}$ depth. One of the objects is selected, and is placed on a table; the robot pushes the object with a cylindrical finger, followed by a retrieval movement.  A colour firewire camera records the action perpendicularly, while a force sensor registers the reaction forces in synchrony with each photograph taken [\fref{fig:scene}(a)].  This set up allowed us to study the effect of using an automatically calibrated physics based model as model for the material, while using 2D meshes, even though the same principles can be used if we use a 3D point cloud and a 3D mesh.  To generate the initial mesh for the mass-spring model a routine placed the nodes for the masses evenly spaced on a rectangular grid, given the number of squares we wanted per side.  We made experiments with grids of $3 \times 2, 10 \times 5$ and $20 \times 10$ squares.  To alleviate the bias produced by the springs they were placed in a star pattern, thus compensating forces on eight directions.

@@

The traditional equations of movement, estimated with Euler or Verlet show big oscillations and frequently become unstable.  Verlet seems to be badly affected by the collision resolution routines and geometric aids, that drastically affect the values of velocity and acceleration.  On the other hand, the customised first degree equations suggested in \sref{sec:ischeme} produce smooth movements that more closely resemble the ground truth, thus obtaining better marks.  From this group of experiments it was also found that evolutionary learning is as efficient when considering all elastic and plastic terms, as when using only a few of them\footnote{The full list of terms is summarised at the end of \sref{sub:EvolAlg}.}.

\subsection{Results of Automatic Parameter Calibration}
\subsubsection{Performance for the Learned Shape Prediction Model}
The performance of the genetic learning algorithm was evaluated using the customised integration scheme and the mass-spring elastic-plastic model.  At this point no prediction of the force was included, rather, the reaction forces measured with the force sensor were used. The F-score was used as fitness function. The values used for the parameters of \alref{alg:genetic} are summarised in \tref{tab:gen_params}, also the minimum and maximum limits for the values of parameters for the mass-spring model.

\begin{table}[!t]
 \renewcommand{\arraystretch}{1.3}
 \caption{Parameters for the Genetic Algorithm}
 \label{tab:gen_params}
 \centering
\begin{tabular}{cc}
\hline
 \bfseries Parameter & \bfseries Value/Range \\  \hline
 \multicolumn{2}{c}{Genetic Algorithm} \\ \hline
 M & 30 \\
 $\alpha = \beta = \gamma = \zeta$  & 5 \\
 $\sigma_{max}$                     & 1000.0 \\
 $\sigma_{min}$                     & 10.0 \\
 $P(s)$                             & 0.05 \\ \hline
 \multicolumn{2}{c}{Elastic Parameters } \\ \hline
 $k_d$                              & $[0.00001, 2]$ \\
 $K_L$                              & $[0.00001, 10000]$ \\
 $K_A$                              & $[0.00001, 10000]$ \\
 $K_{\varphi}$                      & $[0.00001, 100]$ \\
 $max(F)$                           & $[5, 100000]$ \\ \hline
 \multicolumn{2}{c}{Plastic Parameters } \\ \hline
 $\alpha$                           & $[0.00001, 0.9]$ \\
 $yield$                            & $[0.00001, 1.0]$ \\
 $creep$                            & $[0.00001, 1.0]$ \\ \hline
\end{tabular}

\vspace*{1em}
* Where $P(s)$ is the probability of using a special value, like $0.0$, for the spring constants.
\end{table}
